---
title: 'Election Project Step 1 - Source #5'
author: "Ryan Saraie"
date: "November 29, 2016"
output: html_document
---

This step of the project is to create a data frame that consists of data from six sources. Three sources contain election data, and the other three contain census data. In this segment of the project, I am putting data from Source #5 (2010 census) into the complete data frame.

First, I will load the data from the appropriate sources:

```{r}
DataB01 = read.csv("http://www.stat.berkeley.edu/users/nolan/data/voteProject/census2010/B01003.csv")

## Race information

DataDP02 = read.csv("http://www.stat.berkeley.edu/~nolan/data/voteProject/census2010/DP02.csv")

## Socio data

DataDP03 = read.csv("http://www.stat.berkeley.edu/~nolan/data/voteProject/census2010/DP03.csv")

## Economic data

## It's important to note here that all of the data that has just been downloaded is put out in the form of a data frame.
```

Now that all of the data is loaded, let's check to see if any of these data frames have missing values for any of the present variables. We will sapply each data frame to check for NA values at each row, and sum up all of the ones that are found.

```{r}
sum(sapply(DataB01, function(x) sum(is.na(x))))
sum(sapply(DataDP02, function(x) sum(is.na(x))))
sum(sapply(DataDP03, function(x) sum(is.na(x))))

## All of them have an output of 0, so we don't have to worry about clearing NA values.
```

Now we can begin looking at the different variables that we want to incorporate into the data frame.We'll first take a look at HD01_VD01 column in the B01 data frame to make 2 columns of the total estimated population by each county and the estimated white population. The process is that we will take the columns of the county names, signifiers of race, and population values of each race; we make a separate data frame from those columns (titled CensusDF), which will be the main data frame for putting in information.

For reference, the variable names can be found at http://www.stat.berkeley.edu/~nolan/data/voteProject/census2010/B01_metadata.txt 

```{r}
sum(DataB01$POPGROUP.display.label == "Total population")

## This function gives us a good sense of how many counties we're dealing with in total. We have 3217 counties.

CensusDF <- cbind.data.frame(DataB01$GEO.display.label, DataB01$POPGROUP.display.label, DataB01$HD01_VD01) 

## Create a DF of all the columns we need. In this case, we just need the county names, the population labels, and the estimated total population. 

colnames(CensusDF) <- c("County", "POPGROUP.display.label", "Estimated Population")

## Changing the column names to make data editing easier.

## At this point, we have a clean data frame that has all of the variables we want.

```

The issue with CensusDF at this point is that the data frame does not give us separate columns for the white population and for the total population. We do this by creating 2 separate data frames from CensusDF for the total population and the white population; we then combine the population number columns from each to make a data frame that has the county, the white population, and the total population.

```{r}
CensusDFtotal <- CensusDF[CensusDF$POPGROUP.display.label == "Total population", ]
CensusDFwhite <- CensusDF[CensusDF$POPGROUP.display.label == "White alone", ]

## Subsetting the data frames to only give values that display the total population of each county and the white population, which is what we want. 

CensusDFtotal$POPGROUP.display.label <- NULL
CensusDFwhite$POPGROUP.display.label <- NULL

## We remove the POPGROUP.display.label column, as we do not need that information anymore to maintain the clarity of the new data frame.

colnames(CensusDFwhite) <- c("County", "White Population")
colnames(CensusDFtotal) <- c("County", "Total Population")

## We change the column names here to make the data more clear.

CensusDF <- merge(CensusDFwhite, CensusDFtotal, by = "County", all.x = TRUE, all.y = TRUE)

## Merging the two data frames into one.
```

Now we have two columns done, with the estimated total population and the estimated white population. 
We make our next few variable observations by merging columns from the B01, DP02, and DP03 data frames into CensusDF, which will give us a data frame with about 30-40 variables.

Adding several extra columns from the DP02 table, representing different variables by county. We are specifically observing these variables from the 2010 census:

- HC01_VC03 - Estimated # of households
- HC01_VC04 - Estimated # of family households (families)
- HC01_VC06 - Estimated # of family households (families) with children under 18 years
- HC01_VC07 - Estimated # of households with a married-couple family
- HC03_VC08 - Percentage of married families with children under 18 years
- HC03_VC09 - Percentage of families with a male householder and no wife
- HC03_VC11 - Percentage of families with a female householder and no husband
- HC01_VC13 - Estimated # of nonfamily households
- HC03_VC14 - Percentage of households with a lone householder
- HC03_VC15 - Percentage of households with a lone householder at least 65 years old
- HC03_VC17 - Percentage of households with one or more people under 18 years
- HC03_VC18 - Percentage of households with one or more people over 65 years
- HC01_VC20 - Estimated # of average household size
- HC01_VC21 - Estimated # of average family size
- HC03_VC36 - Percentage of people never married
- HC03_VC37 - Percentage of people now married, except separated
- HC03_VC38 - Percentage of people separated

For reference, the variable names can be found at
http://www.stat.berkeley.edu/~nolan/data/voteProject/census2010/DP02_metadata.txt

```{r}
CensusDF <- merge(CensusDF, DataDP02[ ,c("GEO.display.label", "HC01_VC03", "HC01_VC04","HC01_VC06", "HC01_VC07","HC03_VC08", "HC03_VC09", "HC03_VC11", "HC01_VC13", "HC03_VC14", "HC03_VC15", "HC03_VC17", "HC03_VC18", "HC01_VC20", "HC01_VC21", "HC03_VC36", "HC03_VC37", "HC03_VC38")], by.x = "County", by.y = "GEO.display.label", all.x = TRUE, all.y = TRUE)

## We merge CensusDF with all of the relevant DP02 columns, choosing to accept NA values that may be present to allow the data frames to merge.
```

Adding several extra columns from the DP03 table, representing different varialbes by county. We are specifically observing these variables from the 2010 census:

- HC01_VC04 - Estimated # of population 16 years and over with an employment status
- HC03_VC05 - Percentage of population in the labor force
- HC01_VC06 - Estimated # of population in the civilian labor force
- HC03_VC07 - Percentage of people employed in the civilian labor force
- HC03_VC08 - Percentage of people unemployed in the civilian labor force
- HC01_VC10 - Estimated # of population not in the labor force
- HC03_VC13 - Percentage of population unemployed
- HC03_VC16 - Percentage of females 16 and over in the labor force
- HC01_VC17 - Estimated # of females 16 and over in the civilian labor force
- HC03_VC18 - Percentage of females 16 and over employed in the civilian labor force
- HC01_VC31 - Estimated # of people commuting to work by public transit (excluding taxi)
- HC01_VC34 - Estimated # of people working at home
- HC03_VC40 - Estimated # of civilian employed population 16 years and over
- HC01_VC42 - Estimated # of employed civilians with service occupations
- HC01_VC51 - Estimated # of employed civilians in construction
- HC01_VC52 - Estiamted # of employed civilians in manufacturing
- HC01_VC54 - Estimated # of employed civilians in retail trade

For reference, the variable names can be found at
http://www.stat.berkeley.edu/~nolan/data/voteProject/census2010/DP03_metadata.txt

```{r}
CensusDF <- merge(CensusDF, DataDP03[ ,c("GEO.display.label", "HC01_VC04", "HC03_VC05", "HC01_VC06", "HC03_VC07", "HC03_VC08", "HC01_VC10", "HC03_VC13", "HC03_VC16", "HC01_VC17", "HC03_VC18", "HC01_VC31", "HC01_VC34", "HC03_VC40", "HC01_VC42", "HC01_VC51", "HC01_VC52", "HC01_VC54")], by.x = "County", by.y = "GEO.display.label", all.x = TRUE, all.y = TRUE)

## Same process as with the DataDP02 data frame, but with the DP03 data.
```

Now that the entire data frame has all of the data we want, we can begin changing the column names to give some context to each column. We use the colnames function:

```{r}
colnames(CensusDF) <- c("County", "White Population", "Total Population", "Households", "Family Households", "Family Households with Children under 18 years", "Households with Married Couple Family", "% of Married Families with Children under 18 years", "% of Families with Male & no Wife", "% of Families with Wife & no Husband", "Nonfamily Households", "% of Households with Lone Householder", "% of Households with Lone Householder at least 65 years old", "% of Households with at least one under 18 years", "% of Households with at least one over 65 years", "Average Household Size", "Average Family Size", "% of People Never Married", "% of People Now Married, except Separated", "% of People Separated", "Population 16 years and over", "% of Population in Labor Force", "Population in Civilian Labor Force", "% of Population Employed in Civilian Labor Force", "% of Population Unemployed in Civilian Labor Force", "Population not in Labor Force", "% of Population Unemployed", "% of Females 16 years and over in Labor Force", "Females 16 years and over in Civilian Labor Force", "% of Females Employed in Civilian Labor Force", "Population Commuting to Work (Excluding Taxi)", "Population Working at Home", "Employed Population 16 years and over", "Employed with Service Occupations", "Employed in Construction", "Employed in Manufacturing", "Employed in Retail Trade")
```

I now begin the process of merging my data from CensusDF into the larger dataframe. To do this, I separate the state from the county in CensusDF$County, edit the names of the state in CensusDF so that they match up with the main data frame, "results12and16", and then merge the two data frames such that I get all of the info on one table.

```{r}
CensusDF$County <- as.character(CensusDF$County)
list <- strsplit(CensusDF$County, ", ")

## We change CensusDF$County to a character class so we can use strsplit to divide the column by county and state. The presence of a comma in each row makes this process very simple.

library(plyr)
df <- ldply(list)

## We use plyr to make a data frame from our list, which we named "list". The substitute data frame is named "df".

colnames(df) <- c("County", "State")

## Renaming for simplicity.

CensusDF <- cbind(df, CensusDF)

## Combining the data frames.

CensusDF[3] <- NULL

## Removing the original "County" column, which we do not need anymore.

CensusDF$State <- tolower(CensusDF$State)

## Making the entire state column lowercase. The state column in results12and16 is lowercase, so we do this to make the final merge possible.

CensusDF$State <- gsub(" ", "-", CensusDF$State)

## Replacing every space with a dash. As before, we're just making the state columns from the difference data frames identical.

results12and16 <- merge(results12and16, CensusDF, by.x = c("county_name", "state"), by.y = c("County", "State"), all.x = TRUE, all.y = TRUE)

## The final merge.
```

I have now successfully merged all of my data to the master data frame! 